Stuff

### 2014-01-27 paper/Background.tex
paper

Figure \ref{req-resp} shows the data-types corresponding to a request issued
from the processor to the memory subsystem and a response issued from the
memory sub-system to the processor. A request contains the
address corresponding to the request, and a type indicating whether the request
is a load or a store. In case of a store request, a real system will contain the
data to be stored. We augment the request message with two special fields: a
\mylabel which uniquely identifies each request, and an \myindex which
is a natural number denoting the position of a request amongst those emanating
from a particular processor. We omit the data field because it can be subsumed
by the unique label (the data and the unique label together forms another
unique label).

A response issued contains a label to identify the request corresponding to the
response and the processor to which the response is sent to. In the case of a
load request, the memory sub-system returns \stl, the label of the last store
request to that address. In case there has been no store to that address, a
special value called Initial is returned. The memory sub-system also returns a
natural number, \mytime which represents the position at which this request has
been handled globally among all the requests issued from all the processors.
Both the store label (or Initial) and the time are needed just for the
definition of the store-atomicity property -- a real system will simply return
the data corresponding to the load request's address.

\begin{figure}
\centering
\begin{tabular}{|p{.02\textwidth}lp{.9\textwidth}|}
\hline
\multicolumn{3}{|c|}{\Request}\\
\hline
&\proc:& Processor issuing the request\\
&\mylabel:& A unique label to identify the request\\
&\loc:& Address associated with the request\\
&\desc:& Request type (\Ld{} or \St)\\
&\myindex:& A unique natural number indicating the local order of requests\\
\hline
\hline
\multicolumn{3}{|c|}{\Response}\\
\hline
&\proc:& Processor which receives the response\\
&\mylabel:& Label identifying the request corresponding to this response\\
&\stl:& The label of the last store to that address (or the special value,
       Initial) in case of a load response\\
&\mytime:& A natural number indicating the global order of (processing the) requests\\
\hline
\end{tabular}
\caption{Data types for request and response}
\label{req-resp}
\end{figure}

The \Request set in the following contains only those requests processed by the
memory subsystem and the \Response set contains only the responses generated by
the memory subsystem.

The uniqueness of request labels (globally), which is needed for defining store
atomicity, can be defined as follows:
\begin{defn}
\begin{multline*}
\textit{uniqReqLabels} := \forall q_1, q_2\in\Request, q_1.\mylabel = q_2.\mylabel @-> q_1 = q_2\\
\end{multline*}
\end{defn}

Requests from the same processor to the memory subsystem are sent in the order
of their \myindex field. Thus, is request $r$ is sent to a processor $c$, then
the next request that can be sent to the processor obeys the following
property.
\begin{defn}
\begin{multline*}
\textit{nextReq} r c := 
\end{multline*}
\end{defn}

The responses should obey the following sanity conditions:
\begin{defn} (Each response corresponds to a request.)
\begin{equation*}
\textit{respHasReq} := \forall r\in\Response, \exists q\in\Request, r.\mylabel = q.\mylabel
\end{equation*}
\end{defn}

\begin{defn} (A response comes back to the same processor that sent the corresponding request.)
\begin{multline*}
\textit{respProc} := \forall (r\in\Response) (q\in\Request), r.\mylabel = q.\mylabel \\ @-> r.\proc = q.\proc
\end{multline*}
\newline
\end{defn}

\begin{defn} (The labels in the responses are unique. Since \stl{} field is irrelecant for
store requests, they are checked only for load requests.)
\begin{multline*}
\textit{uniqRespLabels} := \forall (r_1, r_2 \in \Response), r_1.\mylabel = r_2.\mylabel @-> \\
\hspace{1cm} r_1.\proc = r_2.\proc \wedge r_1.\mytime = r_2.\mytime \wedge \hfill \\
\forall (q \in \Response), q.\mylabel = r_1.\mylabel @->
q.\desc = \Ld @-> r_1.\stl = r_2.\stl
\end{multline*}
\end{defn}

\begin{defn} (No response can share the same time as another store response.)
\begin{multline*}
\textit{uniqStRespTimes} := 
\forall (r_1, r_2 \in Response), r_1.\mytime = r_2.\mytime @-> \\
\forall (q_1, q_2 \in Request), r_1.\mylabel = q_1.\mylabel @->
r_2.\mylabel = q_2.\mylabel -> \\
q_1.\loc = q_2.\loc @-> q_1.\desc = \St @-> \\ r_1.\mylabel = r_2.\mylabel
\end{multline*}
\end{defn}

In addition, the global ordering of responses to the same processor should not
violate the local ordering of the corresponding requests.
\begin{defn}
\begin{multline*}
\textit{localOrdering} :=
\forall (r_1, r_2 \in \Response) (q_1, q_2 \in \Request), \\ r_1.\mylabel
= q_1.\mylabel @-> r_2.\mylabel = q_2.\mylabel @-> \\ q_1.\proc = q_2.\proc @->
q_1.\loc = q_2.\loc @-> \\ q_1.\myindex < q_2.\myindex @-> r_1.\mytime \le r_2.\mytime
\end{multline*}
\end{defn}

This brings us to the main condition for store atomicity which states that the
store label returned in a load response corresponds to the \emph{last earlier}
store request that has been processed in the system for that address (lastEarlierValue). The
\emph{last} and \emph{earlier} are defined according to the field \mytime{}
returned in the responses. In case no store has happened earlier for a
particular address when a load is processed, then the load response returns a
special value \Initial, which is used to initialize all the addresses (initialValue).

\begin{defn}
\begin{multline*}
\textit{initialValue}:= \forall (r \in \Request) (q \in \Response), r.\mylabel = q.\mylabel @->
q.\desc = \Ld @-> \\(r.\stl = \Initial \wedge
\forall (r' \in \Response) (q' \in \Request),
r'.\mylabel = q'.\mylabel @-> \\ 0 \le r'.\mytime < r.\mytime @->
\neg (q.\loc = q'.\loc \wedge q'.\desc = \St))
\end{multline*}
\end{defn}

\begin{defn}
\begin{multline*}
\textit{lastEarlierValue} := \forall (r \in \Request) (q \in \Response), r.\mylabel = q.\mylabel @-> \\
q.\desc = \Ld @-> 
(\exists m, r.\stl = \Store m \wedge \\
\exists (r_m \in \Response) (q_m \in \Request), 
r_m.\mylabel = m \wedge q_m.\mylabel = m \wedge \\
r_m.\mytime < r.\mytime \wedge q_m.\loc = q.\loc \wedge q_m.\desc = \St \wedge  \\
\forall (r' \in \Response) (q' \in \Request),
r'.\mylabel = q'.\mylabel @-> \\ r_m.\mytime < r'.\mytime < r.\mytime ->
\neg (q.\loc = q'.\loc \wedge q'.\desc = \St))
\end{multline*}
\end{defn}

Using the above definitions, we can define the store-atomicity theorem as follows:
\begin{thm}
\begin{multline*}
storeAtomicity := \textit{uniqReqLabels} @-> \textit{uniqIndices} @-> \\
\textit{respHasReq} \wedge \textit{respProc} \wedge \textit{uniqRespLabels} \wedge
\textit{uniqStRespTimes} \wedge \textit{localOrdering}\\ \wedge 
(\textit{initialValue} \vee \textit{lastEarlierValue})
\end{multline*}
\end{thm}

>>>>

The requests issued by the processors to the memory subsystem can be modeled
using a function \reqFn. $\reqFn(p, i)$ returns the $i^{th}$ request issued by
processor $p$, \ie $\reqFn: (\textit{Processor} * \mathbb{N}) @-> \Request$.
The processor name and the position of the request within the processor
uniquely identifies the request.  Figure \ref{req-resp} shows the data-type
corresponding to a request issued from the processor to the memory subsystem.
A request contains the address (\addrQ) corresponding to the request, and a
type indicating whether the request is a load or a store (\type). In case of a
store request, it also contains data (\dataQ) supplied by the processor for
that address.

Figure \ref{req-resp} also shows the data-type corresponding to a response
issued from the memory subsystem to the processor.  A response contains a label
(\labelR) to identify the request corresponding to the response. The label
contains a processor name and a position, which uniquely identifies a request.
In the case of a load request, the memory sub-system returns \dataR, the data
returned by the memory subsystem for that address. The memory sub-system also
returns a natural number, \timeR, which is used solely in the definition of
store atomicity, and is not returned in a real system. \timeR can be informally
thought to represent the position at which this request has been handled
globally among all the requests issued from all the processors.

\begin{figure}
\centering
\begin{tabular}{|p{.02\textwidth}lp{.9\textwidth}|}
\hline
\multicolumn{3}{|c|}{\Request}\\
\hline
&\addrQ:& Address associated with the request\\
&\desc:& Request type (\Ld{} or \St)\\
&\dataQ:& Data associated with a \St{} request\\
\hline
\hline
\multicolumn{3}{|c|}{\Response}\\
\hline
&\labelR:& Label identifying the request corresponding to this response\\
&\dataR:& Data returned by the memory subsystem for a \Ld{} request\\
&\timeR:& Natural number representing global order of (processing) requests\\
\hline
\end{tabular}
\caption{Data types for request and response}
\label{req-resp}
\end{figure}

In the following, a \Response set only contains those responses sent by the
memory subsystem.

The responses should obey the following sanity conditions:

\begin{defn} (The labels in the responses are unique. Since \dataR{} field is
irrelevant for store requests, so uniqueness of responses does not depend on
\dataR{} field for stores.)
\begin{multline*}
\textit{uniqRespLabels} := \forall (r_1, r_2 \in \Response),\\
r_1.\labelR = r_2.\labelR @-> r_1.\timeR = r_2.\timeR \wedge\\
(\reqFn(r_1.\labelR) = \Ld @-> r_1.\dataR = r_2.\dataR)
\end{multline*}
\end{defn}

\begin{defn} (No response can share the same time as another store response.)
\begin{multline*}
\textit{uniqStRespTimes} := 
\forall (r_1, r_2 \in \Response),
r_1.\timeR = r_2.\timeR @-> \\
(\reqFn (r_1.\labelR)).\addrQ = (\reqFn (r_2.\labelR)).\addrQ @->\\
(\reqFn (r_1.\labelR)).\desc = \St @->
r_1.\labelR = r_2.\labelR
\end{multline*}
\end{defn}

In addition, the global ordering of responses to the same processor should not
violate the local ordering of the corresponding requests.
\begin{defn}
\begin{multline*}
\textit{localOrdering} :=
\forall (r_1, r_2 \in \Response), (r_1.\labelR).\fst = (r_2.\labelR).\fst @->\\
(r_1.\labelR).\snd < (r2.\labelR).\snd @-> r_1.\timeR \le r_2.\timeR
\end{multline*}
\end{defn}

This brings us to the main condition for store atomicity which states that the
value returned for a load response corresponds to the \emph{last earlier} store
request that has been processed in the system for that address
(\textit{lastEarlierValue}). The \emph{last} and \emph{earlier} are defined
according to the field \timeR{} returned in the responses. In case no store has
happened earlier for a particular address when a load is processed, then the
load response returns the initial value for the address (\textit{initialValue}).
\begin{defn}
\begin{multline*}
\textit{initialValue} (r: \Response) := \mylet q := \reqFn (r.\labelR) \myin
q.\desc = \Ld @-> \\ r.\dataR = \initData (r.\addrQ) \wedge \\
\forall (r' \in \Response), \mylet q' := \reqFn (r'.\labelR) \myin 0 \le r'.\timeR < r.\timeR @->\\
\neg (q'.\addrQ = q.\addrQ \wedge q'.\desc = \St)
\end{multline*}
\end{defn}

\begin{defn}
\begin{multline*}
\textit{lastEarlierValue} (r: \Response) := \mylet q := \reqFn (r.\labelR) \myin
q.\desc = \Ld @-> \\ \exists r_m, \mylet q_m := \reqFn (r_m.\labelR) \myin
r.\dataR = q_m.\dataQ \wedge \\
\forall (r' \in \Response), \mylet q' := \reqFn (r'.\labelR) \myin \\ r_m.\timeR \le r'.\timeR < r.\timeR @->
\neg (q'.\addrQ = q.\addrQ \wedge q'.\desc = \St)
\end{multline*}
\end{defn}

Using the above definitions, we can define the store-atomicity theorem as follows:
\begin{thm}
\begin{multline*}
storeAtomicity := 
\textit{uniqRespLabels} \wedge
\textit{uniqStRespTimes} \wedge \textit{localOrdering}\\ \wedge 
(\forall (r \in \Response), \textit{initialValue}(r) \vee \textit{lastEarlierValue}(r))
\end{multline*}
\end{thm}
